# Chronology Statistics

:::{.callout-note title="Partly written"}
The sections below were moved here from A First Pass and have not been revised.

Still planned:

- rbar, EPS and SSS in more depth. These come before Chronology Building, which uses them for cutoffs, stripping and variance stabilization.
- Running rbar and EPS with `rwi.stats.running` as a main section. This output is widely used and often misunderstood.
- `treeMean`: averaging cores to a tree-level series, as used in ecological work where the tree-to-tree differences are the interest.
- `common.interval`, if it fits without being clunky.
:::

## Big Idea

## Packages

We use `dplR` and the on-board Campito Mountain ring widths, `ca533`, detrended as in the Detrending chapter.

```{r}
#| label: stats-setup
library(dplR)
data(ca533)
ca533.rwi <- detrend(rwl = ca533, method = "AgeDepSpline")
```

## Descriptive Statistics for Detrended Data

It is also easy in `dplR` to compute commonly used descriptive statistics that describe the correlation between series (both within and between tree correlations) as well as the expressed population signal, signal-to-noise ratio, and subsample signal strength for a data set. These are done in `dplR` using the `rwi.stats` function so-named because these statistics are typically (but not always) carried out on detrended and standardized ring widths (rwi). If a data set has more than one core taken per tree this information can be used in the calculations to calculate within vs. between tree correlation. The function `read.ids` is used to identify which trees have multiple cores. 

```{r}
ca533.ids <- read.ids(ca533, stc = c(3, 2, 1))
rwi.stats(ca533.rwi, ca533.ids, prewhiten=TRUE)
```

There is (at least) one other way of looking at the average interseries correlation of a data set. The `interseries.cor` function in `dplR` gives a measure of average interseries correlation that is different from the rbar statistics from `rwi.stats`. In this function, correlations are calculated serially between each tree-ring series and a master chronology built from all the other series in the `rwl` object (leave-one-out principle). The average of those correlations is sometimes called the "overall interseries correlation" or even the "COFECHA correlation" in reference to commonly used crossdating software COFECHA. This number is typically higher than the various rbar values given by  `rwi.stats`. We are showing just the first five series and the mean for all series here:

```{r}
ca533.rho <- interseries.cor(ca533.rwi, prewhiten=TRUE,
                             method="spearman")
ca533.rho[1:5, ]
mean(ca533.rho[, 1])
```

Again, if these concepts are unknown to you statistically look at some of the canonical works in dendrochronology like @Cook1990, @Fritts2001, and @Hughes2011.

## Mean Sensitivity

There are a few other descriptive stats that are available. Mean sensitivity (`sens1` and `sens2`) used to be included in the standard summary statistics but **mean sensitivity is actually a terrible statistic that should rarely, if ever, be used [@Bunn2013].** It is at best equivalent to the standard deviation of a time series in cases without high autocorrelation and is an inefficient estimator of the coefficient of variation. So use cautiously if at all.

## Wrapping Up

## Exercises

## Further Reading
